# Retrospective analysis summary for  my.syndromic

```{r}
#loading and preparing the data to work with
load("my.syndromic.RData")
require(vetsyn)
matrix.week <- x@observed
frequency=52
```

## Abortion
```{r}
#create time series
s=1
week    <-  matrix.week[,s]
week.ts <-  ts(week, start = c(x@dates$year[1], x@dates$week[1]),
               frequency = frequency)
t = 1:length(week)

```

```{r fig.width=10, fig.height=5}
#Plot series
plot(week.ts , xlab="Weeks",main="Weeks")
```

### Summary statistics
```{r}
#Percentiles
quantile(week,probs=c(0,0.25,0.5,0.75,1),names=FALSE)
round(mean(week),4)

#Number of weeks at minimum value
(countInMin <- length(which(week == min(week))))
(percentInMin <- round(((countInMin)/(length(week)))*100,2))
```

```{r fig.width=10, fig.height=5}
#ACF and PACF
acf(week,main="ACF for weekly data",lag.max=(frequency+5))
pacf(week,main="PACF for weekly data",lag.max=(frequency+5))
```

### Crude (visual) assessment of temporal effects
```{r fig.width=10, fig.height=5}
boxplot(week ~ x@dates$week, main="Week of the year")
boxplot(week ~ x@dates$year, main = "Year")
```

### POISSON  Regression with sin/cos wave
```{r}
distribution="poisson"

cos = cos (2*pi*t/frequency)
sin = sin (2*pi*t/frequency)
year.f <- as.factor (x@dates$year)
tminus1<-c(week[1],week[1:(length(week)-1)])
tminus2<-c(week[1:2],week[1:(length(week)-2)])
tminus3<-c(week[1:3],week[1:(length(week)-3)])
tminus4<-c(week[1:4],week[1:(length(week)-4)])

fit1 = glm(week ~ year.f, family=distribution)
fit1AR1 = glm(week~ year.f + tminus1, family=distribution)
fit1AR2 = glm(week~ year.f + tminus1+ tminus2, family=distribution)
fit1AR3 = glm(week~ year.f + tminus1+ tminus2 + tminus3, family=distribution)
fit1AR4 = glm(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit2 = glm(week ~ sin + cos, family=distribution)
fit2AR1 = glm(week~ sin + cos + tminus1, family=distribution)
fit2AR2 = glm(week~ sin + cos + tminus1+ tminus2, family=distribution)
fit2AR3 = glm(week~ sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit2AR4 = glm(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit3 = glm(week ~ t + sin + cos, family=distribution)
fit3AR1 = glm(week~ t + sin + cos + tminus1, family=distribution)
fit3AR2 = glm(week~ t + sin + cos + tminus1+ tminus2, family=distribution)
fit3AR3 = glm(week~ t + sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit3AR4 = glm(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

```


```{r}
#Printing AICs
AR1 <- c(fit1=fit1$aic,fit1AR1=fit1AR1$aic,fit1AR2=fit1AR2$aic,fit1AR3=fit1AR3$aic,
                          fit1AR4=fit1AR4$aic)
AR2 <- c(fit2=fit2$aic,fit2AR1=fit2AR1$aic,fit2AR2=fit2AR2$aic,fit2AR3=fit2AR3$aic,
                          fit2AR4=fit2AR4$aic)
AR3 <- c(fit3=fit3$aic,fit3AR1=fit3AR1$aic,fit3AR2=fit3AR2$aic,fit3AR3=fit3AR3$aic,
                          fit3AR4=fit3AR4$aic)


print(AR1)
print(AR2)
print(AR3)

```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Poisson regression")
lines(fit1$fit, col="red"   , lwd=2)
lines(fit2$fit, col="blue"  , lwd=2)
lines(fit3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

### Negative Binomial Regression with sin/cos wave
```{r}
require(MASS)
fitNB1 = glm.nb(week ~ year.f)
fitNB1AR1 = glm.nb(week~ year.f + tminus1)
fitNB1AR2 = glm.nb(week~ year.f + tminus1+ tminus2)
fitNB1AR3 = glm.nb(week~ year.f + tminus1+ tminus2 + tminus3)
fitNB1AR4 = glm.nb(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB2 = glm.nb(week ~ sin + cos)
fitNB2AR1 = glm.nb(week~ sin + cos + tminus1)
fitNB2AR2 = glm.nb(week~ sin + cos + tminus1+ tminus2)
fitNB2AR3 = glm.nb(week~ sin + cos + tminus1+ tminus2 + tminus3)
fitNB2AR4 = glm.nb(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB3 = glm.nb(week ~ t + sin + cos)
fitNB3AR1 = glm.nb(week~ t + sin + cos + tminus1)
fitNB3AR2 = glm.nb(week~ t + sin + cos + tminus1+ tminus2)
fitNB3AR3 = glm.nb(week~ t + sin + cos + tminus1+ tminus2 + tminus3)
fitNB3AR4 = glm.nb(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

```


```{r}
#Printing AICs
AR_NB1 <- c(fitNB1=fitNB1$aic,fitNB1AR1=fitNB1AR1$aic,fitNB1AR2=fitNB1AR2$aic,fitNB1AR3=fitNB1AR3$aic,
                          fitNB1AR4=fitNB1AR4$aic)
AR_NB2 <- c(fitNB2=fitNB2$aic,fitNB2AR1=fitNB2AR1$aic,fitNB2AR2=fitNB2AR2$aic,fitNB2AR3=fitNB2AR3$aic,
                          fitNB2AR4=fitNB2AR4$aic)
AR_NB3 <- c(fitNB3=fitNB3$aic,fitNB3AR1=fitNB3AR1$aic,fitNB3AR2=fitNB3AR2$aic,fitNB3AR3=fitNB3AR3$aic,
                          fitNB3AR4=fitNB3AR4$aic)
print(AR_NB1)
print(AR_NB2)
print(AR_NB3)
```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Negative Binomial Regression")
lines(fitNB1$fit, col="red"   , lwd=2)
lines(fitNB2$fit, col="blue"  , lwd=2)
lines(fitNB3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

```


## GIT
```{r}
#create time series
s=2
week    <-  matrix.week[,s]
week.ts <-  ts(week, start = c(x@dates$year[1], x@dates$week[1]),
               frequency = frequency)
t = 1:length(week)

```

```{r fig.width=10, fig.height=5}
#Plot series
plot(week.ts , xlab="Weeks",main="Weeks")
```

### Summary statistics
```{r}
#Percentiles
quantile(week,probs=c(0,0.25,0.5,0.75,1),names=FALSE)
round(mean(week),4)

#Number of weeks at minimum value
(countInMin <- length(which(week == min(week))))
(percentInMin <- round(((countInMin)/(length(week)))*100,2))
```

```{r fig.width=10, fig.height=5}
#ACF and PACF
acf(week,main="ACF for weekly data",lag.max=(frequency+5))
pacf(week,main="PACF for weekly data",lag.max=(frequency+5))
```

### Crude (visual) assessment of temporal effects
```{r fig.width=10, fig.height=5}
boxplot(week ~ x@dates$week, main="Week of the year")
boxplot(week ~ x@dates$year, main = "Year")
```

### POISSON  Regression with sin/cos wave
```{r}
distribution="poisson"

cos = cos (2*pi*t/frequency)
sin = sin (2*pi*t/frequency)
year.f <- as.factor (x@dates$year)
tminus1<-c(week[1],week[1:(length(week)-1)])
tminus2<-c(week[1:2],week[1:(length(week)-2)])
tminus3<-c(week[1:3],week[1:(length(week)-3)])
tminus4<-c(week[1:4],week[1:(length(week)-4)])

fit1 = glm(week ~ year.f, family=distribution)
fit1AR1 = glm(week~ year.f + tminus1, family=distribution)
fit1AR2 = glm(week~ year.f + tminus1+ tminus2, family=distribution)
fit1AR3 = glm(week~ year.f + tminus1+ tminus2 + tminus3, family=distribution)
fit1AR4 = glm(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit2 = glm(week ~ sin + cos, family=distribution)
fit2AR1 = glm(week~ sin + cos + tminus1, family=distribution)
fit2AR2 = glm(week~ sin + cos + tminus1+ tminus2, family=distribution)
fit2AR3 = glm(week~ sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit2AR4 = glm(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit3 = glm(week ~ t + sin + cos, family=distribution)
fit3AR1 = glm(week~ t + sin + cos + tminus1, family=distribution)
fit3AR2 = glm(week~ t + sin + cos + tminus1+ tminus2, family=distribution)
fit3AR3 = glm(week~ t + sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit3AR4 = glm(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

```


```{r}
#Printing AICs
AR1 <- c(fit1=fit1$aic,fit1AR1=fit1AR1$aic,fit1AR2=fit1AR2$aic,fit1AR3=fit1AR3$aic,
                          fit1AR4=fit1AR4$aic)
AR2 <- c(fit2=fit2$aic,fit2AR1=fit2AR1$aic,fit2AR2=fit2AR2$aic,fit2AR3=fit2AR3$aic,
                          fit2AR4=fit2AR4$aic)
AR3 <- c(fit3=fit3$aic,fit3AR1=fit3AR1$aic,fit3AR2=fit3AR2$aic,fit3AR3=fit3AR3$aic,
                          fit3AR4=fit3AR4$aic)


print(AR1)
print(AR2)
print(AR3)

```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Poisson regression")
lines(fit1$fit, col="red"   , lwd=2)
lines(fit2$fit, col="blue"  , lwd=2)
lines(fit3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

### Negative Binomial Regression with sin/cos wave
```{r}
require(MASS)
fitNB1 = glm.nb(week ~ year.f)
fitNB1AR1 = glm.nb(week~ year.f + tminus1)
fitNB1AR2 = glm.nb(week~ year.f + tminus1+ tminus2)
fitNB1AR3 = glm.nb(week~ year.f + tminus1+ tminus2 + tminus3)
fitNB1AR4 = glm.nb(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB2 = glm.nb(week ~ sin + cos)
fitNB2AR1 = glm.nb(week~ sin + cos + tminus1)
fitNB2AR2 = glm.nb(week~ sin + cos + tminus1+ tminus2)
fitNB2AR3 = glm.nb(week~ sin + cos + tminus1+ tminus2 + tminus3)
fitNB2AR4 = glm.nb(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB3 = glm.nb(week ~ t + sin + cos)
fitNB3AR1 = glm.nb(week~ t + sin + cos + tminus1)
fitNB3AR2 = glm.nb(week~ t + sin + cos + tminus1+ tminus2)
fitNB3AR3 = glm.nb(week~ t + sin + cos + tminus1+ tminus2 + tminus3)
fitNB3AR4 = glm.nb(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

```


```{r}
#Printing AICs
AR_NB1 <- c(fitNB1=fitNB1$aic,fitNB1AR1=fitNB1AR1$aic,fitNB1AR2=fitNB1AR2$aic,fitNB1AR3=fitNB1AR3$aic,
                          fitNB1AR4=fitNB1AR4$aic)
AR_NB2 <- c(fitNB2=fitNB2$aic,fitNB2AR1=fitNB2AR1$aic,fitNB2AR2=fitNB2AR2$aic,fitNB2AR3=fitNB2AR3$aic,
                          fitNB2AR4=fitNB2AR4$aic)
AR_NB3 <- c(fitNB3=fitNB3$aic,fitNB3AR1=fitNB3AR1$aic,fitNB3AR2=fitNB3AR2$aic,fitNB3AR3=fitNB3AR3$aic,
                          fitNB3AR4=fitNB3AR4$aic)
print(AR_NB1)
print(AR_NB2)
print(AR_NB3)
```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Negative Binomial Regression")
lines(fitNB1$fit, col="red"   , lwd=2)
lines(fitNB2$fit, col="blue"  , lwd=2)
lines(fitNB3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

```


## Musculoskeletal
```{r}
#create time series
s=3
week    <-  matrix.week[,s]
week.ts <-  ts(week, start = c(x@dates$year[1], x@dates$week[1]),
               frequency = frequency)
t = 1:length(week)

```

```{r fig.width=10, fig.height=5}
#Plot series
plot(week.ts , xlab="Weeks",main="Weeks")
```

### Summary statistics
```{r}
#Percentiles
quantile(week,probs=c(0,0.25,0.5,0.75,1),names=FALSE)
round(mean(week),4)

#Number of weeks at minimum value
(countInMin <- length(which(week == min(week))))
(percentInMin <- round(((countInMin)/(length(week)))*100,2))
```

```{r fig.width=10, fig.height=5}
#ACF and PACF
acf(week,main="ACF for weekly data",lag.max=(frequency+5))
pacf(week,main="PACF for weekly data",lag.max=(frequency+5))
```

### Crude (visual) assessment of temporal effects
```{r fig.width=10, fig.height=5}
boxplot(week ~ x@dates$week, main="Week of the year")
boxplot(week ~ x@dates$year, main = "Year")
```

### POISSON  Regression with sin/cos wave
```{r}
distribution="poisson"

cos = cos (2*pi*t/frequency)
sin = sin (2*pi*t/frequency)
year.f <- as.factor (x@dates$year)
tminus1<-c(week[1],week[1:(length(week)-1)])
tminus2<-c(week[1:2],week[1:(length(week)-2)])
tminus3<-c(week[1:3],week[1:(length(week)-3)])
tminus4<-c(week[1:4],week[1:(length(week)-4)])

fit1 = glm(week ~ year.f, family=distribution)
fit1AR1 = glm(week~ year.f + tminus1, family=distribution)
fit1AR2 = glm(week~ year.f + tminus1+ tminus2, family=distribution)
fit1AR3 = glm(week~ year.f + tminus1+ tminus2 + tminus3, family=distribution)
fit1AR4 = glm(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit2 = glm(week ~ sin + cos, family=distribution)
fit2AR1 = glm(week~ sin + cos + tminus1, family=distribution)
fit2AR2 = glm(week~ sin + cos + tminus1+ tminus2, family=distribution)
fit2AR3 = glm(week~ sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit2AR4 = glm(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit3 = glm(week ~ t + sin + cos, family=distribution)
fit3AR1 = glm(week~ t + sin + cos + tminus1, family=distribution)
fit3AR2 = glm(week~ t + sin + cos + tminus1+ tminus2, family=distribution)
fit3AR3 = glm(week~ t + sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit3AR4 = glm(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

```


```{r}
#Printing AICs
AR1 <- c(fit1=fit1$aic,fit1AR1=fit1AR1$aic,fit1AR2=fit1AR2$aic,fit1AR3=fit1AR3$aic,
                          fit1AR4=fit1AR4$aic)
AR2 <- c(fit2=fit2$aic,fit2AR1=fit2AR1$aic,fit2AR2=fit2AR2$aic,fit2AR3=fit2AR3$aic,
                          fit2AR4=fit2AR4$aic)
AR3 <- c(fit3=fit3$aic,fit3AR1=fit3AR1$aic,fit3AR2=fit3AR2$aic,fit3AR3=fit3AR3$aic,
                          fit3AR4=fit3AR4$aic)


print(AR1)
print(AR2)
print(AR3)

```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Poisson regression")
lines(fit1$fit, col="red"   , lwd=2)
lines(fit2$fit, col="blue"  , lwd=2)
lines(fit3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

### Negative Binomial Regression with sin/cos wave
```{r}
require(MASS)
fitNB1 = glm.nb(week ~ year.f)
fitNB1AR1 = glm.nb(week~ year.f + tminus1)
fitNB1AR2 = glm.nb(week~ year.f + tminus1+ tminus2)
fitNB1AR3 = glm.nb(week~ year.f + tminus1+ tminus2 + tminus3)
fitNB1AR4 = glm.nb(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB2 = glm.nb(week ~ sin + cos)
fitNB2AR1 = glm.nb(week~ sin + cos + tminus1)
fitNB2AR2 = glm.nb(week~ sin + cos + tminus1+ tminus2)
fitNB2AR3 = glm.nb(week~ sin + cos + tminus1+ tminus2 + tminus3)
fitNB2AR4 = glm.nb(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB3 = glm.nb(week ~ t + sin + cos)
fitNB3AR1 = glm.nb(week~ t + sin + cos + tminus1)
fitNB3AR2 = glm.nb(week~ t + sin + cos + tminus1+ tminus2)
fitNB3AR3 = glm.nb(week~ t + sin + cos + tminus1+ tminus2 + tminus3)
fitNB3AR4 = glm.nb(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

```


```{r}
#Printing AICs
AR_NB1 <- c(fitNB1=fitNB1$aic,fitNB1AR1=fitNB1AR1$aic,fitNB1AR2=fitNB1AR2$aic,fitNB1AR3=fitNB1AR3$aic,
                          fitNB1AR4=fitNB1AR4$aic)
AR_NB2 <- c(fitNB2=fitNB2$aic,fitNB2AR1=fitNB2AR1$aic,fitNB2AR2=fitNB2AR2$aic,fitNB2AR3=fitNB2AR3$aic,
                          fitNB2AR4=fitNB2AR4$aic)
AR_NB3 <- c(fitNB3=fitNB3$aic,fitNB3AR1=fitNB3AR1$aic,fitNB3AR2=fitNB3AR2$aic,fitNB3AR3=fitNB3AR3$aic,
                          fitNB3AR4=fitNB3AR4$aic)
print(AR_NB1)
print(AR_NB2)
print(AR_NB3)
```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Negative Binomial Regression")
lines(fitNB1$fit, col="red"   , lwd=2)
lines(fitNB2$fit, col="blue"  , lwd=2)
lines(fitNB3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

```


## Respiratory
```{r}
#create time series
s=4
week    <-  matrix.week[,s]
week.ts <-  ts(week, start = c(x@dates$year[1], x@dates$week[1]),
               frequency = frequency)
t = 1:length(week)

```

```{r fig.width=10, fig.height=5}
#Plot series
plot(week.ts , xlab="Weeks",main="Weeks")
```

### Summary statistics
```{r}
#Percentiles
quantile(week,probs=c(0,0.25,0.5,0.75,1),names=FALSE)
round(mean(week),4)

#Number of weeks at minimum value
(countInMin <- length(which(week == min(week))))
(percentInMin <- round(((countInMin)/(length(week)))*100,2))
```

```{r fig.width=10, fig.height=5}
#ACF and PACF
acf(week,main="ACF for weekly data",lag.max=(frequency+5))
pacf(week,main="PACF for weekly data",lag.max=(frequency+5))
```

### Crude (visual) assessment of temporal effects
```{r fig.width=10, fig.height=5}
boxplot(week ~ x@dates$week, main="Week of the year")
boxplot(week ~ x@dates$year, main = "Year")
```

### POISSON  Regression with sin/cos wave
```{r}
distribution="poisson"

cos = cos (2*pi*t/frequency)
sin = sin (2*pi*t/frequency)
year.f <- as.factor (x@dates$year)
tminus1<-c(week[1],week[1:(length(week)-1)])
tminus2<-c(week[1:2],week[1:(length(week)-2)])
tminus3<-c(week[1:3],week[1:(length(week)-3)])
tminus4<-c(week[1:4],week[1:(length(week)-4)])

fit1 = glm(week ~ year.f, family=distribution)
fit1AR1 = glm(week~ year.f + tminus1, family=distribution)
fit1AR2 = glm(week~ year.f + tminus1+ tminus2, family=distribution)
fit1AR3 = glm(week~ year.f + tminus1+ tminus2 + tminus3, family=distribution)
fit1AR4 = glm(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit2 = glm(week ~ sin + cos, family=distribution)
fit2AR1 = glm(week~ sin + cos + tminus1, family=distribution)
fit2AR2 = glm(week~ sin + cos + tminus1+ tminus2, family=distribution)
fit2AR3 = glm(week~ sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit2AR4 = glm(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit3 = glm(week ~ t + sin + cos, family=distribution)
fit3AR1 = glm(week~ t + sin + cos + tminus1, family=distribution)
fit3AR2 = glm(week~ t + sin + cos + tminus1+ tminus2, family=distribution)
fit3AR3 = glm(week~ t + sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit3AR4 = glm(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

```


```{r}
#Printing AICs
AR1 <- c(fit1=fit1$aic,fit1AR1=fit1AR1$aic,fit1AR2=fit1AR2$aic,fit1AR3=fit1AR3$aic,
                          fit1AR4=fit1AR4$aic)
AR2 <- c(fit2=fit2$aic,fit2AR1=fit2AR1$aic,fit2AR2=fit2AR2$aic,fit2AR3=fit2AR3$aic,
                          fit2AR4=fit2AR4$aic)
AR3 <- c(fit3=fit3$aic,fit3AR1=fit3AR1$aic,fit3AR2=fit3AR2$aic,fit3AR3=fit3AR3$aic,
                          fit3AR4=fit3AR4$aic)


print(AR1)
print(AR2)
print(AR3)

```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Poisson regression")
lines(fit1$fit, col="red"   , lwd=2)
lines(fit2$fit, col="blue"  , lwd=2)
lines(fit3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

### Negative Binomial Regression with sin/cos wave
```{r}
require(MASS)
fitNB1 = glm.nb(week ~ year.f)
fitNB1AR1 = glm.nb(week~ year.f + tminus1)
fitNB1AR2 = glm.nb(week~ year.f + tminus1+ tminus2)
fitNB1AR3 = glm.nb(week~ year.f + tminus1+ tminus2 + tminus3)
fitNB1AR4 = glm.nb(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB2 = glm.nb(week ~ sin + cos)
fitNB2AR1 = glm.nb(week~ sin + cos + tminus1)
fitNB2AR2 = glm.nb(week~ sin + cos + tminus1+ tminus2)
fitNB2AR3 = glm.nb(week~ sin + cos + tminus1+ tminus2 + tminus3)
fitNB2AR4 = glm.nb(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB3 = glm.nb(week ~ t + sin + cos)
fitNB3AR1 = glm.nb(week~ t + sin + cos + tminus1)
fitNB3AR2 = glm.nb(week~ t + sin + cos + tminus1+ tminus2)
fitNB3AR3 = glm.nb(week~ t + sin + cos + tminus1+ tminus2 + tminus3)
fitNB3AR4 = glm.nb(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

```


```{r}
#Printing AICs
AR_NB1 <- c(fitNB1=fitNB1$aic,fitNB1AR1=fitNB1AR1$aic,fitNB1AR2=fitNB1AR2$aic,fitNB1AR3=fitNB1AR3$aic,
                          fitNB1AR4=fitNB1AR4$aic)
AR_NB2 <- c(fitNB2=fitNB2$aic,fitNB2AR1=fitNB2AR1$aic,fitNB2AR2=fitNB2AR2$aic,fitNB2AR3=fitNB2AR3$aic,
                          fitNB2AR4=fitNB2AR4$aic)
AR_NB3 <- c(fitNB3=fitNB3$aic,fitNB3AR1=fitNB3AR1$aic,fitNB3AR2=fitNB3AR2$aic,fitNB3AR3=fitNB3AR3$aic,
                          fitNB3AR4=fitNB3AR4$aic)
print(AR_NB1)
print(AR_NB2)
print(AR_NB3)
```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Negative Binomial Regression")
lines(fitNB1$fit, col="red"   , lwd=2)
lines(fitNB2$fit, col="blue"  , lwd=2)
lines(fitNB3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

```


## Systemic
```{r}
#create time series
s=5
week    <-  matrix.week[,s]
week.ts <-  ts(week, start = c(x@dates$year[1], x@dates$week[1]),
               frequency = frequency)
t = 1:length(week)

```

```{r fig.width=10, fig.height=5}
#Plot series
plot(week.ts , xlab="Weeks",main="Weeks")
```

### Summary statistics
```{r}
#Percentiles
quantile(week,probs=c(0,0.25,0.5,0.75,1),names=FALSE)
round(mean(week),4)

#Number of weeks at minimum value
(countInMin <- length(which(week == min(week))))
(percentInMin <- round(((countInMin)/(length(week)))*100,2))
```

```{r fig.width=10, fig.height=5}
#ACF and PACF
acf(week,main="ACF for weekly data",lag.max=(frequency+5))
pacf(week,main="PACF for weekly data",lag.max=(frequency+5))
```

### Crude (visual) assessment of temporal effects
```{r fig.width=10, fig.height=5}
boxplot(week ~ x@dates$week, main="Week of the year")
boxplot(week ~ x@dates$year, main = "Year")
```

### POISSON  Regression with sin/cos wave
```{r}
distribution="poisson"

cos = cos (2*pi*t/frequency)
sin = sin (2*pi*t/frequency)
year.f <- as.factor (x@dates$year)
tminus1<-c(week[1],week[1:(length(week)-1)])
tminus2<-c(week[1:2],week[1:(length(week)-2)])
tminus3<-c(week[1:3],week[1:(length(week)-3)])
tminus4<-c(week[1:4],week[1:(length(week)-4)])

fit1 = glm(week ~ year.f, family=distribution)
fit1AR1 = glm(week~ year.f + tminus1, family=distribution)
fit1AR2 = glm(week~ year.f + tminus1+ tminus2, family=distribution)
fit1AR3 = glm(week~ year.f + tminus1+ tminus2 + tminus3, family=distribution)
fit1AR4 = glm(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit2 = glm(week ~ sin + cos, family=distribution)
fit2AR1 = glm(week~ sin + cos + tminus1, family=distribution)
fit2AR2 = glm(week~ sin + cos + tminus1+ tminus2, family=distribution)
fit2AR3 = glm(week~ sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit2AR4 = glm(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

fit3 = glm(week ~ t + sin + cos, family=distribution)
fit3AR1 = glm(week~ t + sin + cos + tminus1, family=distribution)
fit3AR2 = glm(week~ t + sin + cos + tminus1+ tminus2, family=distribution)
fit3AR3 = glm(week~ t + sin + cos + tminus1+ tminus2 + tminus3, family=distribution)
fit3AR4 = glm(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4, family=distribution)

```


```{r}
#Printing AICs
AR1 <- c(fit1=fit1$aic,fit1AR1=fit1AR1$aic,fit1AR2=fit1AR2$aic,fit1AR3=fit1AR3$aic,
                          fit1AR4=fit1AR4$aic)
AR2 <- c(fit2=fit2$aic,fit2AR1=fit2AR1$aic,fit2AR2=fit2AR2$aic,fit2AR3=fit2AR3$aic,
                          fit2AR4=fit2AR4$aic)
AR3 <- c(fit3=fit3$aic,fit3AR1=fit3AR1$aic,fit3AR2=fit3AR2$aic,fit3AR3=fit3AR3$aic,
                          fit3AR4=fit3AR4$aic)


print(AR1)
print(AR2)
print(AR3)

```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Poisson regression")
lines(fit1$fit, col="red"   , lwd=2)
lines(fit2$fit, col="blue"  , lwd=2)
lines(fit3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

### Negative Binomial Regression with sin/cos wave
```{r}
require(MASS)
fitNB1 = glm.nb(week ~ year.f)
fitNB1AR1 = glm.nb(week~ year.f + tminus1)
fitNB1AR2 = glm.nb(week~ year.f + tminus1+ tminus2)
fitNB1AR3 = glm.nb(week~ year.f + tminus1+ tminus2 + tminus3)
fitNB1AR4 = glm.nb(week~ year.f  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB2 = glm.nb(week ~ sin + cos)
fitNB2AR1 = glm.nb(week~ sin + cos + tminus1)
fitNB2AR2 = glm.nb(week~ sin + cos + tminus1+ tminus2)
fitNB2AR3 = glm.nb(week~ sin + cos + tminus1+ tminus2 + tminus3)
fitNB2AR4 = glm.nb(week~ sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

fitNB3 = glm.nb(week ~ t + sin + cos)
fitNB3AR1 = glm.nb(week~ t + sin + cos + tminus1)
fitNB3AR2 = glm.nb(week~ t + sin + cos + tminus1+ tminus2)
fitNB3AR3 = glm.nb(week~ t + sin + cos + tminus1+ tminus2 + tminus3)
fitNB3AR4 = glm.nb(week~ t + sin + cos  +
                tminus1+ tminus2+ tminus3+ tminus4)

```


```{r}
#Printing AICs
AR_NB1 <- c(fitNB1=fitNB1$aic,fitNB1AR1=fitNB1AR1$aic,fitNB1AR2=fitNB1AR2$aic,fitNB1AR3=fitNB1AR3$aic,
                          fitNB1AR4=fitNB1AR4$aic)
AR_NB2 <- c(fitNB2=fitNB2$aic,fitNB2AR1=fitNB2AR1$aic,fitNB2AR2=fitNB2AR2$aic,fitNB2AR3=fitNB2AR3$aic,
                          fitNB2AR4=fitNB2AR4$aic)
AR_NB3 <- c(fitNB3=fitNB3$aic,fitNB3AR1=fitNB3AR1$aic,fitNB3AR2=fitNB3AR2$aic,fitNB3AR3=fitNB3AR3$aic,
                          fitNB3AR4=fitNB3AR4$aic)
print(AR_NB1)
print(AR_NB2)
print(AR_NB3)
```

```{r fig.width=10, fig.height=5}
plot(t,week, type="l",main="Negative Binomial Regression")
lines(fitNB1$fit, col="red"   , lwd=2)
lines(fitNB2$fit, col="blue"  , lwd=2)
lines(fitNB3$fit, col="green" , lwd=2)
legend("topleft",pch=3,col=c("red","blue","green"),
                  c("year", "sin/cos","t + sin/cos"))

```


